---
title: "Class17"
author: "Olivia Solorzano"
format: pdf
editor: visual
---

Pertussis, or whooping cough, is a highly contagious lung infection cause by a bacteria *B. pertussis*

The CDC tracks reported cases in the U.S. since the 1920s.

```{r}
cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
         Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )
```


> Q1.

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(Year, Cases) +
  geom_point() +
  geom_line() +
  labs(x="Year", y="Cases")
```

The first big "whole-cell" pertussi vaccine program started in 1942

> Q2. 

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(Year, Cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept=1946, col="blue") + 
  geom_vline(xintercept=1996, col="red", linetype=2)
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

We could possibly have a better system of detection of the disease, parents may not vaccinate children because they think that vaccinations for TDaP and other vaccines bring neurological diseases, or the bacterium became stronger and the vaccine is no longer effective to destroy it.  


Something bi is happening witgh pertussis cases and big outbreaks are once againa a major public health concern!  BUGGER

One of the main hypothesis for the increasing case numbers is waning vaccine efficacy with the newer aP vaccine. 

Enter the CMI-PB project, which is studying this problem on large scale. Let's see what data they have.

Their data is available in JSON format (key: value pair style). We will use the "jsonlight" package to read their data. 

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

>Q5. How many Male and Female subjects/patients are in the dataset?

>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$biological_sex)
table(subject$race)
table(subject$race, subject$biological_sex)
```

Let's read more database tables from CMI-PB: 

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", 
                      simplifyVector = TRUE)
head(specimen, 3)
```


I want to "join" (a.k.a "merge"/link/etc.) the `subject` and `specimen` tables together. I will use the **dyplr** package for this. 
```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```

```{r}
ab <- read_json("https://www.cmi-pb.org/api/ab_titer", 
                      simplifyVector = TRUE)
```

Now I can join the "meta" that we made above and contains all info about the subjects and specimens with this `ab` data. 
```{r}
abdata <- inner_join(meta, ab)
head(abdata)
```

```{r}
dim(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```
> Q12. What do you notice about the number of visit 8 specimens compared to other visits?

```{r}
table(abdata$visit)
```
There are way less vsiit 8 specimens because the project is still ongoing and we have not got that data for all individuals yet.

##Examine IgG1 Ab titer levels

We will use the filter() function from dplyr to focus on just IgG1 isotype and visits 1 to 7 (i.e. examine our )

```{r}
ig1 <- filter(abdata, isotype == "IgG1", visit!=8)
head(ig1)
```

```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot() + 
  facet_wrap(vars(visit), nrow=2)
```
> Q14. What antigens show differences in the level of IgG1 antibody titers recognizing them over time? Why these and not others?

FIM2/3 show differences in MFI. This is "Fimbrial protein" that makes the backteria pilus and is involved in cell adhesion.

PT Pertosuss Toxin
FHA Filementous Hemagluttinin

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  theme_bw()
```

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“Measles”, that is not in our vaccines) and a clear antigen of interest (“FIM2/3”, extra-cellular fimbriae proteins from B. pertussis that participate in substrate attachment).

```{r}
filter(ig1, antigen=="Measles") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```


```{r}
filter(ig1, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```
FIM2/3 MFI levels rise more in comparison to Measles antigen in both aP and wP subjects. 





