---
title: "Class06Hw"
format: pdf
editor: visual
---

Example input vectors to start with 

```{r}
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

Average grade for student 1.
```{r}
mean(student1)
```


Average grade for student 2. 
`is.na()` function is used to convert all *NA* inputs of student 2 grades into 0
```{r}
student2[is.na(student2)] <- 0
student2
mean(student2)
```

Average grade for student 3
`is.na()` function is used to convert all *NA* inputs of student 2 grades into 0
```{r}
student3[is.na(student3)] <- 0
student3
mean(student3)
```

Q1. Average grade with lowest score dropped for student 1. I can use the minus sign together with `which.min()` to exclude the lowest value: 
```{r}
student1
which.min(student1)
student1[ - which.min(student1)]
```


Now I need to put this all back together to make our working snippet: 
```{r}

#Map/Replace NA values to zero
student3 [ is.na(student3) ]  <- 0 
#Exclude the lowest score
student3 [ -which.min(student3)]
# Calculate the mean
mean(student3[ -which.min(student3)] )


```


This is my working snippet that I can turn into a function called `grade()`

All function in R have at leat 3 things:
- **Name**, in our case "grade"
- Input **arguments**, student 1 etc.
- **Body**, this is our working snippet above. 

```{r}
grade <- function(y) {
    #Map/Replace NA values to zero
    y [ is.na(y) ]  <- 0 
    #Exclude the lowest score
    y [ -which.min(y)]
    # Calculate the mean
    mean(y[ -which.min(y)] )
}

```

Using the function to get student 1 grade:
```{r}
grade(student1)
```


Read a gradebook from online
```{r}
hw <- read.csv("https://tinyurl.com/gradeinput", row.names=1)
hw
```


> Q1. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score.

The `apply` functions allows us to run any function over the rows or columns of data.frame. 
```{r}
ans <- apply(hw, 1, grade)
ans
```


> Q2. Using your grade() function and the supplied gradebook, who is the top scoring student overall in the gradebook? 

```{r}
ans [ which.max(ans) ]
```


> Q3. From your analysis of the gradebook, which homework was toughest on students?

```{r}
which.min( apply (hw, 2, sum, na.rm=TRUE) )
```
Hw 2 seems to gave been the toughest for students.



> Q4. Optional Extension: From your analysis of the gradebook, which homework was most
predictive of overall score (i.e. highest correlation with average grade score)?

```{r}
hw$hw1
```

```{r}
ans
```

```{r}
cor(hw$hw1, ans)
cor(hw$hw3, ans)
```

Changing NA values in homework assignments to 0.
```{r}
mask <- hw 
mask [ is.na(mask)] <- 0
mask 
```

```{r}
cor(mask$hw5, ans)
```


```{r}
  apply(mask, 2, cor, y=ans)
```










