---
title: "Class 10 : Halloween Candy"
format: pdf
editor: visual
---
In today's class we will examine 538 Candy data and see if this helps us gain some more feeling for how PCA and other methods work. 

```{r}
candy <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv", row.names = 1)
head(candy)
```


> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy[,2])
```

> Q. What are these fruity candy?

We can use the == 
```{r}
rownames( candy[ candy$fruity == 1, ] )

```


>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

## How often does my favorite candy win

```{r}
candy["Kit Kat",]$winpercent
```


>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the `winpercent` column is on a 0:100 scale and all others appear to be on a 0:1 scale.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero here means the candy is not classified as containing chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy, aes(winpercent, )) +
  geom_histogram(bins=10)

#or

hist(candy$winpercent)
```


>Q9. Is the distribution of winpercent values symmetrical?

No

> Q10. Is the center of the distribution above or below 50%?

Below 50% with a mean: 

```{r}
mean(candy$winpercent)
median(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

To answer this question I will need to: 
- "subset" (a.k.a. "select", "filter") the candy dataset to just chocolate candy
- Get their winpercent values
- Calculate the mean of these
- Do the same for fruity and compare. 

```{r}
# Filter/select/subset to just chocolate rows
chocolate.candy <-  candy[ as.logical(candy$chocolate), ]

# Get their winpercent values
chocolate.winpercent <- chocolate.candy$winpercent

# Calculate mean of chocolate winpercent
mean( chocolate.winpercent )


```


```{r}
# Filter/select/subset to just fruity rows
fruity.candy <- candy[ as.logical(candy$fruity), ]

# Get their winpercent values
fruity.winpercent <- fruity.candy$winpercent

# Calculate mean of fruity winpercent
mean( fruity.winpercent )
```

Chocolate candy is on a higher ranked than fruity candy. 

> Q12. Is this difference statistically significant?

The p-value is lower than 0.05 and thus shows that the results are statistically significant and people prefer chocolate. 

```{r}
t.test(chocolate.winpercent, fruity.winpercent)
```


There is a function called `sort()` for sorting vectors of input

```{r}
x <- c(5, 2, 10)

# sort(x, decreasing = TRUE)
sort(x)

```

The buddy function to `sort()` that is often more useful is called `order()`. It returns the "indices" of the input that would result in it being sorted. 

```{r}
order(x)
x[ order(x)]
```


> Q13. What are the five least liked candy types in this set?

```{r}
ord <- order(candy$winpercent)
ord
head(candy[ord,])
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
ord_decreasing <- order(candy$winpercent, decreasing = TRUE)
ord_decreasing
head(candy[ord_decreasing,])
```


> Q15. Make a first barplot of candy ranking based on winpercent values.

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
#To make it more colorful
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```


```{r}

library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent))  +
  geom_col(fill = my_cols)
```

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst

## Taking a look at pricepercent

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?


```{r}
ggplot(candy) +
  aes(winpercent, pricepercent) +
  geom_point(col=my_cols)
```

Let's add labels
```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text(size=3)
```
To deal with overlapping lables I can use the **ggrepel** package.

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(size = 2, max.overlaps = 8)
```

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```



# 5 Exploring the correlation structure

Pearson correlation goes between -1 and +1 with zero indicating no correlation and values close to one being very highly (ani) correlated. 

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are anti-correlated.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent are most positively correlated.

# 6. Principal Component Analysis

The base R function for PCA is called `prcomp()` and we can set the "scale = TRUE/FALSE"

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```


The main result of PCA - i.e. the new OC plot (projection of candy on our new PC axis) is contained in `pca$x`

```{r}
pc <- as.data.frame(pca$x)

ggplot(pc) +
  aes(PC1, PC2) +
  geom_point(col=my_cols)
  #geom_text_repel(max.overlaps = 5)
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
The variables fruity, hard, and pluribus are in the positive direction of PC1 which make sense because all of these are positively correlated.

